Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-26 demystify.tex
updates

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for \emph{international students}, but has been expanded to include information for \emph{US domestic students} (\autoref{sec:domestic-students}).
Moreover, while this primarily aims at students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other good R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

\begin{mybox}
This document is open source and updated regularly. Its latest version is available at

@@

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.
Much of these uncertainties stem from the lack of information and guidance on the admission process in the US (social places like Facebook and Reddit are full of confusion and contradictory information).  This document aims to address these concerns.

@@

\section{Who is this document for?}
\sectioninfo{This document is intended for international students from smaller countries and domestic students from less well-known universities.}

While this guideline can be useful for all students, it would most benefit \textbf{international students} from \textbf{smaller countries and less well-known universities} (or domestic students from US universities with no PhD programs or limited research opportunities).

@@

This document also aim to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc. There are numerous \href{https://github.com/nguyenthanhvuh/phd-cs-us}{resources online} from students  that tell you \emph{``what''} to do, but few explain the \emph{``why''}. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

While this document may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions, which are frankly the majority of good schools in the US.

@@

While being ambitious and aiming high is good, you should understand how PhD admission works (e.g., by reading \emph{this} document and realizing things such as having a good GPA or GRE doesn't mean much to top programs), that in the US there are many good schools, and just be realistic.

@@

Most of what is written in this document applies to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university. Different universities might have different definitions, e.g., permanent residents or US citizens.} and international students.  However, there are some differences and benefits that domestic students should be aware of and can leverage to improve their chances of admission.
\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships and funding for domestic students).
That makes you stand out and can help your case.

@@

%\autoref{tab:ranking} lists .
\chapter{History and Acknowledgement}\label{sec:ack}
\paragraph{History} This document was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about PhD programs in the US.

I was also a member of the large VietPhD group on Facebook and often browse Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most participants are students in non-CS fields or not in the US, and their answers are not always accurate and can lead to more confusion. So I thought it would be useful to have a document that is specific to CS PhD programs in the US from an insider perspective.

I started writing this document in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The document was initially intended for international students but has been generalized to all students.
The document is open-source and available on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.
\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (Mason). Before Mason, I was at the University of Nebraska-Lincoln (UNL). Both UNL and Mason are large R1 research universities in the US and have many international students in their CS graduate programs. My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

@@

Though I was not an international student, many of my students and collaborators are/were. I also mentor students from Vietnam and have close colleagues and friends who were once international students. I hope to capture the diverse challenges and experiences they've faced in this document so that it can be a valuable resource for prospective international students.
Finally, my upbringing in the US provides a perspective aligned with American culture, allowing me to shed light on various issues, particularly those related to cultural differences (\autoref{sec:cultural}).

@@

Also thanks to NSF for encouraging faculty to be creative in research and education, which allows me to write this document.

>>>>

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this book was intended for \emph{international students}, but has been expanded to include information for \emph{US domestic students} (\autoref{sec:domestic-students}).
Moreover, while this primarily aims at students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this book know about, the information should be generalizable to other good R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

\begin{mybox}
This book is open source and updated regularly. Its latest version is available at

@@

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.
Much of these uncertainties stem from the lack of information and guidance on the admission process in the US (social places like Facebook and Reddit are full of confusion and contradictory information).  This book aims to address these concerns.

@@

\section{Who is this handbook for?}
\sectioninfo{This book is intended for international students from smaller countries and domestic students from less well-known universities.}

While this book can be useful for all students, it would most benefit \textbf{international students} from \textbf{smaller countries and less well-known universities} (or domestic students from US universities with no PhD programs or limited research opportunities).

@@

This book also aim to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc. There are numerous \href{https://github.com/nguyenthanhvuh/phd-cs-us}{resources online} from students  that tell you \emph{``what''} to do, but few explain the \emph{``why''}. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

While this handbook may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions, which are frankly the majority of good schools in the US.

@@

While being ambitious and aiming high is good, you should understand how PhD admission works (e.g., by reading \emph{this} handbook and realizing things such as having a good GPA or GRE doesn't mean much to top programs), that in the US there are many good schools, and just be realistic.

@@

Most of what is written in this handbook applies to both domestic\footnote{For simplicity domestic means you did your undergrad (or MS) at a US university. Different universities might have different definitions, e.g., permanent residents or US citizens.} and international students.  However, there are some differences and benefits that domestic students should be aware of and can leverage to improve their chances of admission.
\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships and funding for domestic students).
That makes you stand out and can help your case.

@@

%\autoref{tab:ranking} lists .
\chapter{History and Acknowledgement}\label{sec:ack}
\paragraph{History} This handbook was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about PhD programs in the US.

I was also a member of the large VietPhD group on Facebook and often browse Internet forums (e.g., \texttt{Reddit/gradadmission} and \texttt{GradCafe}). There I saw many questions from students about PhD programs.  However, most participants are students in non-CS fields or not in the US, and their answers are not always accurate and can lead to more confusion. So I thought it would be useful to have a handbook that is specific to CS PhD programs in the US from an insider perspective.

I started writing this book in May 2023 and have been updating it since then (mostly around deadline time when I tend to procrastinate, i.e., \emph{productive procrastination}!). The book was initially intended for international students but has been generalized to all students.
The handbook is open-source and available on \href{https://github.com/nguyenthanhvuh/phd-cs-us}{GitHub} so that others can contribute to it.
\paragraph{About Me} I am an associate professor in the CS dept at George Mason University (Mason). Before Mason, I was at the University of Nebraska-Lincoln (UNL). Both UNL and Mason are large R1 research universities in the US and have many international students in their CS graduate programs. My personal and lab website is at \href{https://dynaroars.github.io}{dynaroars.github.io}.

@@

Though I was not an international student, many of my students and collaborators are/were. I also mentor students from Vietnam and have close colleagues and friends who were once international students. I hope to capture the diverse challenges and experiences they've faced in this book so that it can be a valuable resource for prospective international students.
Finally, my upbringing in the US provides a perspective aligned with American culture, allowing me to shed light on various issues, particularly those related to cultural differences (\autoref{sec:cultural}).

@@

Also thanks to NSF for encouraging faculty to be creative in research and education, which allows me to write this handbook.
